---
title: Women in Parliament -- data.table

# Use letters for affiliations
author: 
  - name: Saghir Bashir
#     affiliation: a
#   - name: Second Author
#     affiliation: a,b

# address:
#   - code: a
#     address: Institute of Smoke and Magic, University of Sometown, Sometown, XY, 12345
#   - code: b
#     address: Department of Neat Tricks, Whereever State University, Someplace, MC, 67890
    
# Optional: line of arbitrary text with additional information.
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"
#date_subtitle: Published in *Journal of Statistical Software*, 2018

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: "[CC BY SA](https://creativecommons.org/licenses/by-sa/4.0/) [ilustat](https://ilustat.com/) $\\bullet$ [info@ilustat.com](mailto:info@ilustat.com)"

# Place eg a DOI URL or CRAN Package URL here
doi_footer: "Learn more at https://ilustat.com/resources/"

# Abstract
abstract: |
  We will use the World Bank’s indicator data for “Women in Parliament” as a case 
  study when working with the data.table R package. We will guide you through the 
  geographical and time trends for the percentage of women in national parliaments. 
  We will start by learning about and understanding the raw data, which we will 
  then process (“wrangle”) in preparation for some exploratory analysis.

# Optional: Acknowledgements
# acknowledgements: |
  
# Optional: One or more keywords
keywords:
  - Women in Parliament
  - World Bank Indicator
  - data.table
  - Tinyverse

# Paper size for the document, values of letterpaper and a4paper
papersize: a4

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
numbersections: true

# Optional: Specify the depth of section number, default is 5
secnumdepth: 3

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
bibliography: pinp

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "Women in Parliament -- data.table"

# Produce a pinp document
output: 
  pinp::pinp:
    fig_caption: yes
    collapse: true
    keep_tex: false
    latex_engine: xelatex
    
# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{YourPackage}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE) #, knitr.table.format = 'latex')
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
options(width=45)
knitr::opts_chunk$set(fig.pos = 'H')

# Reset the data.table print options.
options(datatable.print.topn=4, digits=3)

# Finding files using here package
library(here)
here()

# Making some aesthetic changes for this document
theme_set(theme_gray(base_size = 9))
update_geom_defaults("point", list(size = 0.5))
update_geom_defaults("boxplot", list(outlier.size = 0.5))
```

# Preface

We present a real-life case study for the data.table[^dtref] package using the 
World Bank’s “Women in Parliament” indicator data. To get the most out of this 
case-study guide, repeat the examples and do the exercises whilst reading it.

## Guide materials

You can download materials for this guide from this link:

- [https://ilustat.com/shared/WiP-rdatatable.zip](https://ilustat.com/shared/WiP-rdatatable.zip) 

Unzip the file, which contains the data, this guide and an R script exercise file. 
We advise you to work with "`WiP-Exercise.R`" file to follow the examples and do 
the exercises. If you are using RStudio, you can double click on "`WiP-dt.Rproj`" 
to get started. 

# Objectives

*Explore the geographical and time trends for the percentage of women[^objNote] in 
national parliaments.*

# Understanding the Data

## The World Bank Data 

The raw data for *"Proportion of seats held by women in national parliaments"* 
includes the percentage of women in parliament (_"single or lower parliamentary chambers 
only"_) by country (region) and year. It can be downloaded from:[^wbstats]

- https://data.worldbank.org/indicator/SG.GEN.PARL.ZS 

As part of its "open data" mission the World Bank offers _"free and open access 
to global development data"_ kindly licensed under the "Creative Commons Attribution 
4.0 (CC-BY 4.0)".[^CC4]

## Source Data

The data originates from the "Inter-Parliamentary Union" (IPU)[^ipuwebsite] which
provides an *_"Archive of statistical data on the percentage of women in 
national parliaments"_* going back to 1997 on a monthly basis:

- http://archive.ipu.org/wmn-e/classif-arc.htm

The World Bank data is for “single or lower parliamentary chambers only”, while 
the IPU also presents data for “Upper Houses or Senates”. Moreover, the IPU provides 
the actual numbers used to calculate the percentages (which the World Bank does not).

## Data limitations

Take caution when interpreting these data, as parliamentary systems vary from country 
to country, and in some cases over time. Some of the issues to consider include:

- Who has, and who does not have, the right to become a Member of Parliament (MP)?
- How does someone become an MP? Through democratic elections? How is "democratic
election" defined? 
- What is the real power of MPs and their parliament? Can MPs make a difference?

## Data definitions & assumptions

### "Women"

The definition for “women” is not given, so we will assume that it refers to a 
binary classification for gender (sex).

### "Country (Region)"

The definition of countries and regions can change over time.  (e.g. formation of new
countries after conflicts, new member states joining a pre-existing collective). 
How are these changes reflected in the data? How do they affect the interpretation? 

## Pro tip

Understand the limitations of your data before anybody else points them out to you. 

# About the data file

The data is stored in a file called:

- ``r list.files(path = here("data"), pattern = "^API")``

To simplify things we have copied it to `WB-WiP.csv` (which also allows us to 
maintain the original file in case something goes wrong).

## Pro tip

Always keep a backup copy of the data. Alternatively, set the data file(s) to 
“read-only” to protect it from being overwritten or modified.

## Exercise

It is important to look at and understand the contents of the file before you start
using it. Using a text editor or a spreadsheet software, open the `WB-WiP.csv` file
(in the `data` directory). What do you observe in the contents of this file?

## Content and Structure

The first four lines of WB-WiP.csv can be ignored, since they contain two lines 
of meta-information and two blank lines, as follows:
```{bash first4lines, echo=FALSE, comment=NA}
head -n 4 ../data/WB-WiP.csv | cat -n |  sed 's/^[[:blank:]]*/ /g'
```

The fifth line contains the column (variable) names and the body of data starts 
in the sixth line. It is important to note that there was no collection of data 
for a majority of the years, which means that it is “missing”.

# Importing the data

Based on our findings above, we can "skip" the first four lines and treat the
fifth line as column (variable) names. 
```{r readData, collapse=TRUE}
library(data.table)
library(here)
wip <- fread(here("data", "WB-WiP.csv"), 
             skip = 4, header = TRUE)
```

## Exercise

Check what you have read by typing “`wip`” in the console window. What do you 
observe? Type “`class(wip)`” and “`str(wip)`” to confirm that “`wip`” is of class 
“`data.table`”. 

## "Fix" column names

Some of the column names contain spaces while others are numeric:
```{r colNames, collapse=TRUE}
head(names(wip))
tail(names(wip))
```
By using the `make.names()` function we don't need to use back ticks 
(`` ` ``) around the column names (e.g. `` `r '\x60col name\x60'` ``).
```{r fixColNames, collapse=TRUE}
names(wip) <- make.names(names(wip))
head(names(wip))
tail(names(wip))
```

# Data Wrangling Aims

We can simplify the production of summaries and plots by restructuring the current 
`wip` dataset (which has `r ncol(wip)` columns) to the following format:
```
   Country       Year   pctWiP
   Country AAA   1997     ##.#
   Country AAA   1998     ##.#
   Country AAA   1999     ##.#
   ...
```
pctWiP refers to the percentage of women in parliament.

## Key information retained

These three columns will contain the same information as the `wip` dataset but in
a more usable format. We will also add a variable for the ratio of male to female MPs.

## Superfluous columns

We will start by removing columns ``r tail(names(wip), n=1)``, `Indicator.Name` and 
 `Indicator.Code`. There are years without any data but they will be removed 
 automatically later (when restructuring from "wide" to "long" format).

Column ``r tail(names(wip), n=1)`` is created automatically due to an extra comma 
at the end of the column names (fifth) line of `WB-WiP.csv`:
```{bash extraComma, echo=FALSE, comment="   ..."}
head -n 5 ../data/WB-WiP.csv | tail -n -1 -c 31
```
## Check

Before removing it check that all values are `NA`.
```{r checkV64, collapse=TRUE}
wip[, .N, by=.(V64)]
```
Column `Indicator.Name` has the unique value _"`r unique(wip[, .(Indicator.Name)])`"_ 
and in `Indicator.Code` it is _"`r unique(wip[, .(Indicator.Code)])`"_. 
As there is only one indicator in this dataset we will remove these two columns.

## Exercise

Confirm that both `Indicator.Name` and `Indicator.Code` have the same values for
all observations. Hint: Use the approach above for variable ``r tail(names(wip), n=1)``

## Removing columns

The indicator and ``r tail(names(wip), n=1)`` columns can be removed. We will also 
rename "`Country.Name`" as "`Country`" and "`Country.Code`" as "`Code`".
```{r rmCols, collapse=TRUE}
wip[, c("Indicator.Name", "Indicator.Code", 
        "V64"):=NULL]
setnames(wip, c("Country.Name", "Country.Code"), 
              c("Country", "Code"))
head(names(wip))
tail(names(wip))
```

## Reshape to long format

We want to transform the data so that for each country the year (column) data becomes
a row. At the same time we will remove the missing data (with the `na.rm` option).
```{r meltwip, collapse=TRUE, message=FALSE, warning=FALSE}
WP <- melt(wip,
           id.vars = c("Country", "Code"),
           measure = patterns("^X"),
           variable.name = "YearC",
           value.name = c("pctWiP"),
           na.rm = TRUE)
WP
```


## Final tweaks to `WP`

Create a numeric `Year` variable and a `Ratio` of men to women in parliament.
```{r finalTweaks, collapse=TRUE}
WP[, `:=`(Year=as.numeric(gsub("[^[:digit:].]", 
                               "",  YearC)),
          Ratio = (100-pctWiP)/pctWiP)][
            , YearC:=NULL]
setcolorder(WP, c("Country", "Code", "Year", 
                  "pctWiP", "Ratio"))
# Look at the contents of WP
WP
```

# Questions

The objective is to look at the geographical and time trends in the data. We will
answer the following questions.

- What are the time trends for Portugal?
- How does Portugal compare to other countries?
- Which countries have the highest percentage of women in parliament by year?
- How do continents compare?
- What are the global trends over time?

## Exercise - Without Programming

- Which country do you think has the highest percentage of women in parliament? 
- In each continent (i.e. Africa, Americas, Asia, Europe and Oceania), which country
has the highest percentage of women in parliament? 
- What is the world percentage of women in parliament in `r max(WP$Year)`?

# Exploratory Analysis

## Select a country

This guide explores how Portugal performs over time and compared to other countries. 
Before continuing, select another country for yourself to repeat the examples and 
do the exercises.

## Time trends for Portugal

First look at the raw data.
```{r PTTable}
WP[Country %in% "Portugal"]
```

### Visualisation

It is easier to find trends within a plot.
```{r PTplot, fig.width=3, fig.height=2.3}
library(ggplot2)
library(magrittr)
WP[Country %in% "Portugal"] %>% 
ggplot(aes(Year, pctWiP)) +
  geom_line() + geom_point() +
  scale_y_continuous(limits=c(0, 50)) +
  ylab("% Women in Parliament")
```

### Interpretation

In 1990 Portugal had 7.6% women in parliament (i.e. 12.2 men for each woman), which 
increased to 34.8% (i.e. 1.87 men for each woman) in 2018. This still falls short 
of 50% (i.e. point of gender parity in parliament).

## Exercise 

For your chosen country look at the time trend data and the plot. What is your 
interpretation? How does it compare to Portugal?

## Portugal versus European Union (EU) countries

We selected six EU countries (due to space limitations) for comparison. It would 
be better to compare all EU and/or all European countries.
```{r euPctPlot, fig.width=3.5, fig.height=2.9, cache=FALSE}
WP[Country %in% c("Portugal", "Sweden", "Spain",
     "Hungary", "Romania", "Finland", "Germany",
                           "European Union")] %>%
  ggplot(aes(Year, pctWiP, colour=Country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0, 50), 
                     breaks=seq(0, 50, by=10)) +
  ggtitle("Women in Parliament: EU Countries") +
  ylab("% Women in Parliament")
```

### Interpretation

Since 2007 Portugal has had more women in parliament than the European Union average. 
Hungary and Romania both had a higher percentage of women in parliament in 1990 
(around the end of the Cold War) than they have had since. The key point to note 
is that none of these countries reaches equality between males and females in 
parliament, although Sweden and Finland come closest.

## A couple of points to note

### "Germany"

In October 1990, the process of “German reunification” lead to the creation of 
Germany, which united the former “German Democratic Republic” (East Germany) and 
the “Federal Republic of Germany” (West Germany). Therefore, since reunification, 
the data is presented for the reunified “Germany” only. Careful thought should be 
given to handling, analysing and interpreting any pre-reunification data (if available).

### "European Union"

The “European Union” has changed over time (unlike the “continent of Europe”). It 
started in the 1950s as a block of six European countries (known as the “European 
Community”) and has expanded over the years to 28 countries (with the United Kingdom 
about to depart). This raises the question of how the European Union average is 
calculated. For a given year, is it calculated based on the actual member states 
in that year or on all of the current member states?

## Exercises

Compare the country of your choice to four or five other countries by plotting a 
line graph similar to the one above. 

## Countries with the highest percentage of women in parliament

A quick answer can be obtained by looking at the highest percentages.
```{r allTopPct}
WP[order(-pctWiP), head(.SD, 10)]
```

## Data speaks

Are you surprised? Data can be very enlightening.

## Highest percentage by year

Which countries have the highest percentage of women in parliament by year?
```{r allTopPctYear, collapse=TRUE}
WP[order(Year, -pctWiP), head(.SD, 1), by = Year]
```

## Merging continent

The variable `Country` in the `WP` dataset is a mix of countries and regions (e.g. 
"European Union", "South Asia" and "World"). To present the highest percentages 
grouped by continent we need to add it. Luckily, given the large number of R packages 
available, we can merge the "`continent`" from the "`codelist`" dataset in the 
"`countrycode`" package. 
```{r mergeContinent}
# Ensure that 'countrycode' package is installed.
# install.packages("countrycode")
library(countrycode)
cl <- as.data.table(codelist)[, .(continent, wb)]
setnames(cl, c("continent", "wb"), 
             c("Continent", "Code"))
cWP <- cl[WP, on="Code"]
```

## Highest percentages by year and continent

Which countries have the highest percentages in 1990 and 2018?
```{r allTopPctYearContinent, collapse=TRUE}
cWP[Year %in% c(1990, 2018) & !is.na(Continent)][
    order(Year, -pctWiP), head(.SD, 1), 
    by = .(Year, Continent)][
    order(Continent, Year), 
    .(Continent, Year, Country, pctWiP)]
```

## Decline in percentage

Which countries have had a decline in percentage since their first measurement
(not always 1990)?

```{r declinePct, collapse=TRUE}
dWP <- cWP[
  order(Country, Year), .SD[c(1,.N)], 
   by=Country][,
  pctDiff := pctWiP - shift(pctWiP), by=Country][
  pctDiff<0][
  order(pctDiff)]
dWP[!is.na(Continent),
    .(Country, pctWiP, pctDiff)]
```

### Visualisation 

We will plot the trend lines for countries with at least a 5% decline. Note that the
"5%" is arbitrarily selected.

```{r decline5pct, fig.width=3.5, fig.height=2.5}
# Select the countries to plot
dclpct <- unique(dWP[!is.na(Continent) &
                   pctDiff <= -5]$Country)

WP[Country %in% dclpct] %>%
  ggplot(aes(Year, pctWiP, colour=Country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0, 40),
  breaks=seq(0, 40, by=10)) +
  ggtitle("Women in Parliament: Decline >=5%") +
  ylab("% Women in Parliament")
```

### Interpretation

There is a consistent decline between 1990 and 1997 that should be investigated 
in collaboration with a subject matter expert to understand the potential causes.

## Ranked status

Another way to look at the data is to look at the ranking of countries, which 
could be done at a global level or by continent. Nonetheless, the results should 
be interpreted with caution and an understanding of the actual percentages. For 
example, if most countries were around the 50% mark, rankings could be misleading 
and subject to random fluctuations.

## Global ranks by year

We will rank the countries by year based on the percentage of women in parliaments. 
The countries with the highest percentage will be ranked first and the lowest last. 
A total for the number of countries with data is included as it varies by year.
```{r globalRank , collapse=TRUE}
cWP[!is.na(Continent), 
    `:=`(RankG = rank(-pctWiP), TotalG = .N),
        by = .(Year)]
```

## Global ranking -- Portugal

```{r globalRankPT , collapse=TRUE}
cWP[Country=="Portugal", 
  .(Country, Year, pctWiP, Ratio, RankG, TotalG)][
  order(Year)]
```

### Interpretation

Portugal has generally been ranked in the first quartile (25%) of countries in 
the world, with the fluctuations of its ranking most likely due to random variation. 

## Exercise 

For your chosen country, interpret its ranking over the years. How does it compare 
to Portugal?

## Continent ranks by year

We will rank the countries by year within a continent based on the percentage of 
women in parliaments. The countries with the highest percentage will be ranked 
first and the lowest last. A total for the number of countries with data, within 
each continent, is included as it varies by year.
```{r continentRank , collapse=TRUE}
cWP[!is.na(Continent), 
    `:=`(RankC = rank(-pctWiP), TotalC = .N),
        by = .(Continent, Year)]
```
## Portugal's ranking in Europe

```{r continentRankPT , collapse=TRUE}
cWP[Country=="Portugal", 
  .(Country, Year, pctWiP, Ratio, RankC, TotalC)][
  order(Year)]
```

## Plot of Portugal's ranking in Europe

Below we reproduce the percentage plot to show how Portugal ranks in relation to 
six other European countries. Note that the highest percentage is ranked first 
and the lowest last.
```{r euRankplot, fig.width=3.5, fig.height=2.7}
cWP[Country %in% c("Portugal", "Sweden", "Spain",
  "Hungary", "Romania", "Finland", "Germany")] %>%
  ggplot(aes(Year, RankC, colour=Country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0, 45), 
                     breaks=seq(0, 45, by=10)) +
  ggtitle("Women in Parliament: Ranked") +
  ylab("Rank in Europe")
```

### Interpretation

A total of `r nrow(cWP[Continent=="Europe" & Year==1990])` European countries had
data in 1990, `r nrow(cWP[Continent=="Europe" & Year==1997])` in 1997 and 
`r nrow(cWP[Continent=="Europe" & Year==2018])` in 2018. Within Europe, Portugal 
was typically ranked in the second quartile (25-50%) with the fluctuations of its 
ranking most likely due to random variation.

## Exercise 

How does your chosen country rank within its continent?

## Highest rank by year and continent

Which countries have the highest rank in 1990 and 2018? The answer will coincide 
with the highest percentages (see above).

```{r allTopRankYearContinent, collapse=TRUE, echo=2}
options(datatable.print.rownames=FALSE)
cWP[Year %in% c(1990, 2018) & RankC==1][
    order(Continent, Year), 
      .(Continent, Year, Country, pctWiP, RankC)]
options(datatable.print.rownames=TRUE)
```

## Overall picture

What are the trends globally? There are various regions defined in the World Bank
data. We can plot them and highlight the world "average".

```{r globalTrends, warning=FALSE, message=FALSE, fig.width=3, fig.height=2.5}
library(gghighlight)
cWP[is.na(Continent)] %>%
  ggplot(aes(Year, pctWiP, group=Country)) +
  geom_line() +
  gghighlight(Country=="World", 
              use_direct_label = FALSE) +
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0, 40), 
                     breaks=seq(0, 40, by=10)) +
  ggtitle("Women in Parliament: Global Trends") +
  ylab("% Women in Parliament")
```

### Interpretation

The grey lines show that regardless of how we define region the general trends
are upwards. The "World" percentage (black line) increased between 1997 and 2018. 
In 2018, women in parliament represented 24% (i.e. a ratio of 3.17 men to each woman), 
which is still less than half the level before gender parity can be claimed.

# Conclusion

This guide presented an analysis of the percentage of women in parliament as a real-life 
case study for the data.table package. Although the format limited what could be 
presented, we can conclude that the percentage of women in parliament is increasing 
but that gender parity in parliaments is still far-off.

There is a lot more that can be said and discussed about the limitations, interpretation 
and potential impact of this data which the World Bank has nicely [summarised](https://databank.worldbank.org/data/reports.aspx?source=2&type=metadata&series=SG.GEN.PARL.ZS#).[^WBinterpretation] You are strongly encouraged to read their discussion 
for a more complete understanding.



[^dtref]: For more information on the `data.table` package see http://r-datatable.com/.
[^wbstats]: The `wbstats` R package (https://cran.r-project.org/web/packages/wbstats/)
gives access to a  "tidier" version of the World Bank indicator data.
[^objNote]: The objective could be termed neutrally as "gender trends" but we will 
keep it per the World Bank data.
[^CC4]: https://datacatalog.worldbank.org/public-licenses#cc-by.
[^ipuwebsite]:Inter-Parliamentary Union: https://www.ipu.org/.
[^WBinterpretation]: https://databank.worldbank.org/data/reports.aspx?source=2&type=metadata&series=SG.GEN.PARL.ZS.
